Type-Safe Modular Parsing

### 2016-11-29 paper/sections/Overview.tex
Overview

With Packrat parser combinators, we obtain the possibility of writing parsers in a modular way. Let's start by looking at how to build a parser for a simple language, which only has variables and applications at the beginning.

@@

Firstly, we define the abstract syntax by Scala's convenient case classes. Expressions have type \lstinline{Expr}, and each of them is either \lstinline{Var} or \lstinline{App}.

\begin{lstlisting}
abstract class Expr
case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
\end{lstlisting}

And then we write parsing functions for each case.

\begin{lstlisting}
val pVar: PackratParser[Expr] =
  ident ^^ Var

@@

The \lstinline{PackratParser} type and the combinators are imported from the Scala Packrat parsing library. Some details about these structures will be discussed in later sections, so that readers could only focus on the types of these functions at the first reading.

\lstinline{pVar} is straight forward. It parses a single identifier, and its type \lstinline{PackratParser[Expr]} indicates the result is an \lstinline{Expr}. However the type of \lstinline{pApp} is different. It takes a parser of \lstinline{Expr} as an argument, which represents the parser of the whole expression, namely \lstinline{e} here. And use it to parse two consecutive expressions as an application. In this particular case, the parser of the whole expression is in fact the combination of \lstinline{pVar} and \lstinline{pApp}. We can write equivalent code as below.

\begin{lstlisting}
val pApp2: PackratParser[Expr] = {
  lazy val e = pApp2 | pVar
  e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}

\lstinline{pApp2} works fine, but it does not have the extensibility as \lstinline{pApp} does. Since the parser of the whole expression in \lstinline{pApp2} is hard-coded, it covers and only covers the variable and application cases. If we we extend \lstinline{Expr} later to add a third case lambda abstraction, we have to edit the existing code. The parameterization technique used in \lstinline{pApp} is called \textit{open recursion}, which is discussed in section \ref{sec:openandparsing}.

We can use a trait to glue them together, and define \lstinline{pVarApp} as a combination of the two parsing functions. A type synonym \lstinline{Fix} is used to shorten the types.

@@

\begin{lstlisting}
def parseAndPrint(p: Fix[PackratParser[Expr]])(inp: String) = {
  val t = phrase(fix(p))(new lexical.Scanner(inp))
  if (t.successful) println(t.get) else scala.sys.error(t.toString)
}

parseAndPrint(new ExprParser {}.pVarApp)("x y")
\end{lstlisting}

It will print \lstinline{App(Var(x),Var(y))} as the parsing result of string \lstinline{"x y"}.

@@

\begin{lstlisting}
parseAndPrint(new LambdaParser {}.pVarAppLam)("(\\x.x) y")
\end{lstlisting}

The code above will print result \lstinline{App(Lam(x,Var(x)),Var(y))} from parsing \lstinline{"(\\x.x)} \lstinline{y"}.

@@

We glue them together by defining an object.

@@

\begin{lstlisting}
def parse[E](inp: String)(alg: VarApp.Alg[E]): E = {
  val p = new VarApp.Parser[E] {}.pVarApp(alg)
  val t = phrase(fix(p))(new lexical.Scanner(inp))
  if (t.successful) t.get else scala.sys.error(t.toString)
}

def parseAndPrint(inp: String) = println(parse(inp)(new VarApp.Print{}))

parserAndPrint("x y")
\end{lstlisting}

>>>>

In this section we start demonstrating how to write modular parsers. Suppose we want to parse a simple language of variables and applications.

@@

The abstract syntax can be defined by Scala's case classes, and then we write parsers for the two cases \lstinline{Var} and \lstinline{App}.

\begin{lstlisting}
abstract class Expr
case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr

val pVar: PackratParser[Expr] = ident ^^ Var

@@

Some structures are imported from the Scala Packrat parsing library, such as the combinator \lstinline{^^}. They will be discussed in later sections, so that readers could only focus on the type of parsers at the first reading.

\lstinline{pVar} for parsing variables is straight forward. It parses a single identifier, and its type \lstinline{PackratParser[Expr]} indicates the result is an \lstinline{Expr}.

The type of \lstinline{pApp} is different. It takes a parser of \lstinline{Expr} as an argument, and use it to parse two consecutive expressions as an application. This argument \lstinline{e} represents the parser of \emph{the whole expression}. In this particular case, it is in fact the combination of \lstinline{pVar} and \lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the code still works. However, using an abstract argument provides extensibility to the parser. Because once the abstract syntax is extended later, the parser of the whole expression will change correspondingly. Then we have to edit existing code if it is hard-coded. The parameterization technique used in \lstinline{pApp} is called \textit{open recursion}, which is discussed in section \ref{sec:openandparsing}.

We can use a trait to glue them together, and define \lstinline{pVarApp} as a combination of the two parsers. A type synonym \lstinline{Fix} is used to shorten the types.

@@

\begin{lstlisting}
def parse[E](p: Fix[PackratParser[E]])(inp: String): E = {
  val t = phrase(fix(p))(new lexical.Scanner(inp))
  if (t.successful) t.get else scala.sys.error(t.toString)
}

println(parse(new ExprParser {}.pVarApp)("x y"))
\end{lstlisting}

It will print the AST as \lstinline{App(Var(x),Var(y))}.

@@

\begin{lstlisting}
println(parse(new LambdaParser {}.pVarAppLam)("(\\x.x) y"))
\end{lstlisting}

The code above will print \lstinline{App(Lam(x,Var(x)),Var(y))} from parsing \lstinline{"(\\x.x)} \lstinline{y"}.

@@

We glue them together by defining an object. It actually becomes a \textit{language component} that represents a specific language feature.

@@

\begin{lstlisting}
def parseWithAlg[E](inp: String)(alg: VarApp.Alg[E]): E = {
  val p = new VarApp.Parser[E] {}.pVarApp(alg)
  parse[E](p)(inp)
}

println(parseWithAlg("x y")(new VarApp.Print{}))
\end{lstlisting}

With Object Algebras, we can extend all the stuff, including abstract syntax, operations, and parsers. The code below extend the language to support lambda abstraction.

\begin{lstlisting}
object Lambda {
  trait Alg[E] extends VarApp.Alg[E] {
    def Lam(x: String, e: E): E
  }

  trait Print extends Alg[String] with VarApp.Print {
    def Lam(x: String, e: String) = "\\" + x + "." + e
  }

  trait Parser[E] extends VarApp.Parser[E] {
    lexical.delimiters += ("\\", ".")

    val pLam: Alg[E] => Fix[PackratParser[E]] = alg => e =>
      ("\\" ~> lcid) ~ ("." ~> e) ^^ { case x ~ e0 => alg.Lam(x, e0) }
    val pVarAppLam: Alg[E] => Fix[PackratParser[E]] =
      pVarApp | pLam
  }
}
\end{lstlisting}

For easily combining the parsers using Object Algebras as the representation of abstract syntax, we have a special combinator \huang{we need to rename our combinator} defined in our library.

In addition to make extending operations possible, another advantage of using Object Algebras is that it supports multiple sorts of syntax. For example, if our language has not only expressions but also types, we can easily distinguish them in the abstract syntax by an extra type parameter. This topic will be discussed in section \ref{subsec:differentsyntax}.

\begin{lstlisting}
trait TypedLambda[E, T] extends VarApp.Alg[E] {
  def Lam(x: String, t: T, e: E): E
  def IntType(): T
  def BoolType(): T
  def ArrowType(a: T, b: T): T
}
\end{lstlisting}
